On joint estimation of ancestral states and phylogenetic model parameters

### 2017-10-09 deprecated.tex
a few ideas trying to work through general case

Our plan of attack is to focus on the likelihood
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
    &=        p_{\emptyset}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
    &\qquad + p_{1}          \cdot\log(1-x'^2+y'^2-x'^2y'^2) \\
    &\qquad + p_{2}          \cdot\log(1+x'^2-y'^2-x'^2y'^2) \\
    &\qquad + p_{3}          \cdot\log(1-x'^2+y'^2-x'^2y'^2) \\
    &\qquad + p_{12}         \cdot\log(1-x'^2-y'^2+x'^2y'^2) \\
    &\qquad + p_{13}         \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
    &\qquad + p_{23}         \cdot\log(1-x'^2-y'^2+x'^2y'^2) \\
    &\qquad + p_{123}        \cdot\log(1+x'^2-y'^2-x'^2y'^2) \\
    &\qquad - \log(8) \\
    &\qquad + p_{\emptyset}  \cdot\log((1+x')^2   (1+w')(1+y')^2) \\
    &\qquad + p_{1}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \\
    &\qquad + p_{2}          \cdot\log((1+x')^2   (1+w')(1+y')(1-y')) \\
    &\qquad + p_{3}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \\
    &\qquad + p_{12}         \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y')) \\
    &\qquad + p_{13}         \cdot\log((1-x')^2   (1+w')(1+y')^2) \\
    &\qquad + p_{23}         \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y')) \\
    &\qquad + p_{123}        \cdot\log((1+x')^2   (1+w')(1+y')(1-y')).
\end{align*}
This simplifies to
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
    &=        p_{\emptyset}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
    &\qquad + p_{13}         \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
    &\qquad + (p_{1} + p_{3})          \cdot\log(1-x'^2+y'^2-x'^2y'^2) \\
    &\qquad + (p_{2} + p_{123})          \cdot\log(1+x'^2-y'^2-x'^2y'^2) \\
    &\qquad + (p_{12} + p_{23})         \cdot\log(1-x'^2-y'^2+x'^2y'^2) \\
    &\qquad + (2-p_{1}-p_{3}-p_{12}-p_{23}-2p_{13})\log(1+x') \\
    &\qquad + (p_{1}+p_{3}+p_{12}+p_{23}+2p_{13})\log(1-x') \\
    &\qquad + (2-p_{2}-p_{12}-p_{23}-p_{123})\log(1+y') \\
    &\qquad + (p_{2}+p_{12}+p_{23}+p_{123})\log(1-y') \\
    &\qquad + \log(1+w').
\end{align*}
This likelihood occurs when $\hat{w}' = 1$ and $\hat{y}' > \hat{x}'$, the latter coming from
$$
(1-x'^2)(1+y'^2) > (1+x'^2)(1-y'^2)
$$
only if $y' > x'$.
We simply need to obtain
$$
(\hat{x}', \hat{y}', \hat{w}') = \arg\max_{x, y, w} \ \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
$$
as functions of $x$ and $y$ and verify the region where $\hat{w}'(x, y) = 1$ and $\hat{y}'(x, y) > \hat{x}'(x, y)$.
This region will be consistent with one always estimating $\hat{w}'$ to be one.
This can similarly be done for the other likelihood term with $\hat{x}'(x, y) > \hat{y}'(x, y)$.

@@

>>>>

Our plan of attack is to focus on the likelihood
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
    &=        p_{\emptyset}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
    &\qquad + p_{1}          \cdot\log(1-x'^2+y'^2-x'^2y'^2) \\
    &\qquad + p_{2}          \cdot\log(1+x'^2-y'^2-x'^2y'^2) \\
    &\qquad + p_{3}          \cdot\log(1-x'^2+y'^2-x'^2y'^2) \\
    &\qquad + p_{12}         \cdot\log(1-x'^2-y'^2+x'^2y'^2) \\
    &\qquad + p_{13}         \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
    &\qquad + p_{23}         \cdot\log(1-x'^2-y'^2+x'^2y'^2) \\
    &\qquad + p_{123}        \cdot\log(1+x'^2-y'^2-x'^2y'^2) \\
    &\qquad - \log(8) \\
    &\qquad + p_{\emptyset}  \cdot\log((1+x')^2   (1+w')(1+y')^2) \\
    &\qquad + p_{1}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \\
    &\qquad + p_{2}          \cdot\log((1+x')^2   (1+w')(1+y')(1-y')) \\
    &\qquad + p_{3}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \\
    &\qquad + p_{12}         \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y')) \\
    &\qquad + p_{13}         \cdot\log((1-x')^2   (1+w')(1+y')^2) \\
    &\qquad + p_{23}         \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y')) \\
    &\qquad + p_{123}        \cdot\log((1+x')^2   (1+w')(1+y')(1-y')).
\end{align*}
This simplifies to
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
    &=        p_{\emptyset}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
    &\qquad + p_{13}         \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
    &\qquad + (p_{1} + p_{3})          \cdot\log(1-x'^2+y'^2-x'^2y'^2) \\
    &\qquad + (p_{2} + p_{123})          \cdot\log(1+x'^2-y'^2-x'^2y'^2) \\
    &\qquad + (p_{12} + p_{23})         \cdot\log(1-x'^2-y'^2+x'^2y'^2) \\
    &\qquad + (2-p_{1}-p_{3}-p_{12}-p_{23}-2p_{13})\log(1+x') \\
    &\qquad + (p_{1}+p_{3}+p_{12}+p_{23}+2p_{13})\log(1-x') \\
    &\qquad + (2-p_{2}-p_{12}-p_{23}-p_{123})\log(1+y') \\
    &\qquad + (p_{2}+p_{12}+p_{23}+p_{123})\log(1-y') \\
    &\qquad + \log(1+w').
\end{align*}
This likelihood occurs when $\hat{w}' = 1$ and $\hat{y}' > \hat{x}'$, the latter coming from
$$
(1-x'^2)(1+y'^2) > (1+x'^2)(1-y'^2)
$$
only if $y' > x'$.

We simply need to obtain
$$
(\hat{x}', \hat{y}', \hat{w}') = \arg\max_{x, y, w} \ \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
$$
as functions of $x$ and $y$ and verify the region where $\hat{w}'(x, y) = 1$ and $\hat{y}'(x, y) > \hat{x}'(x, y)$.
This region will be consistent with one always estimating $\hat{w}'$ to be one.
This can similarly be done for the other likelihood term with $\hat{x}'(x, y) > \hat{y}'(x, y)$.

@@

\subsubsection{Maximization of functionals---calculus of variations?}

Call the likelihood $\ell$.
The likelihood is maximized at functions of $x$ and $y$ with the property that $\hat{x}'(x,y) = \hat{y}'(y,x)$.
Let's call $\hat{y}'(x, y) := g(x, y)$.
Look at
\begin{align*}
    \ell
    &=        p_{\emptyset}  \cdot\log(1+g(y, x)^2+g(x, y)^2+4g(y, x)g(x, y)w'+g(y, x)^2g(x, y)^2) \\
    &\qquad   p_{13}  \cdot\log(1+g(y, x)^2+g(x, y)^2-4g(y, x)g(x, y)w'+g(y, x)^2g(x, y)^2) \\
    &\qquad + (p_{1} + p_{3})          \cdot\log(1-g(y, x)^2+g(x, y)^2-g(y, x)^2g(x, y)^2) \\
    &\qquad + (p_{2} + p_{123})          \cdot\log(1+g(y, x)^2-g(x, y)^2-g(y, x)^2g(x, y)^2) \\
    &\qquad + (p_{12} + p_{23})         \cdot\log(1-g(y, x)^2-g(x, y)^2+g(y, x)^2g(x, y)^2) \\
    &\qquad + (2-p_{1}-p_{3}-p_{12}-p_{23}-2p_{13})\log(1+g(y, x)) \\
    &\qquad + (p_{1}+p_{3}+p_{12}+p_{23}+2p_{13})\log(1-g(y, x)) \\
    &\qquad + (2-p_{2}-p_{12}-p_{23}-p_{123})\log(1+g(x, y)) \\
    &\qquad + (p_{2}+p_{12}+p_{23}+p_{123})\log(1-g(x, y)) \\
    &\qquad + \log(1+w')
\end{align*}
with the restriction that $g(x, y) > g(y, x)$.
We now have two problems: 1) what to do with $w'$?; and how to solve the two-variable functional maximization?
We may be able to take $\hat{w}'(x, y)$ to be like our quadratic formula version but now with 
$$
\alpha_1 := \alpha_1(x, y) = 1+x^2+y^2+4xy^2+x^2y^2,
$$
$$
\alpha_2 := \alpha_2(x, y) = 1+x^2+y^2-4xy^2+x^2y^2,
$$
$$
\beta := \beta(x, y) = 1+g(y, x)^2+g(x, y)^2+g(y, x)^2g(x, y)^2,
$$
$$
\gamma := \gamma(x, y) = 4g(y, x)g(x, y).
$$
So in a way we want again the region where $\hat{w}'(x, y) \ge 1$ with
$$
\hat{w}'(x, y) = \frac{-b(x, y) - \sqrt{b(x, y)^2 - 4a(x, y)c(x, y)}}{2a(x, y)}.
$$
We have a functional maximization of one two-variable function such that
\begin{align*}
    \ell
    &=        p_{\emptyset}  \cdot\log(1+g(y, x)^2+g(x, y)^2+4g(y, x)g(x, y)\hat{w}'(g)+g(y, x)^2g(x, y)^2) \\
    &\qquad   p_{13}  \cdot\log(1+g(y, x)^2+g(x, y)^2-4g(y, x)g(x, y)\hat{w}'(g)+g(y, x)^2g(x, y)^2) \\
    &\qquad + (p_{1} + p_{3})          \cdot\log(1-g(y, x)^2+g(x, y)^2-g(y, x)^2g(x, y)^2) \\
    &\qquad + (p_{2} + p_{123})          \cdot\log(1+g(y, x)^2-g(x, y)^2-g(y, x)^2g(x, y)^2) \\
    &\qquad + (p_{12} + p_{23})         \cdot\log(1-g(y, x)^2-g(x, y)^2+g(y, x)^2g(x, y)^2) \\
    &\qquad + (2-p_{1}-p_{3}-p_{12}-p_{23}-2p_{13})\log(1+g(y, x)) \\
    &\qquad + (p_{1}+p_{3}+p_{12}+p_{23}+2p_{13})\log(1-g(y, x)) \\
    &\qquad + (2-p_{2}-p_{12}-p_{23}-p_{123})\log(1+g(x, y)) \\
    &\qquad + (p_{2}+p_{12}+p_{23}+p_{123})\log(1-g(x, y)) \\
    &\qquad + \log(1+\hat{w}'(g)).
\end{align*}
